Practical Ethics for Computational Semantics\\ \large Philosophy of Distributional Semantics

### 2017-06-02 pds/Paper/paper_pds.tex
last commit otd

We will now briefly go over the five principles - \emph{responsibility, explainability, accuracy, auditability and fairness} to describe their original intention in the PAA.
Responsibility requires developers to take seriously the possibility that adverse effects are produced their algorithmic system. Consequently, there should be designated roles within the organization to be responsible for redress if harm is caused. Excuses akin to ''the algorithm did it'' are no valid scapegoats. This reflects the acknowledgment that, at some level, there are people who design these systems and put them to use.
Explainability demands that both data selection and decisions taken by the algorithms can be explained to people affected by its usage in terms that are understandable by laywomen. The point here is not to obscure the workings of the systems to a point where no stakeholder can grasp it.
Accuracy asks designers and users of the system to keep track of its mistakes and sources of error, so that worst case scenarios become more easily anticipated and possible damage reduced.
Audibility is the requirement to create transparency in such a way that interested third parties are in the position to review the algorithms performance. This point is related to explainability.
Finally, Fairness requires that the algorithmic system does not affect different demographic groups in disparate and unfavorable ways, that is, discriminates against them.
We now want to assess to what degree we can use these principles for the field of DS.
\section{Principles for Distributional Semantics}\hypertarget{sec4}{}
It should be fair to say that some of these principles allow seamless transfer from general ML to any specific sub-domain, as DS, without much further discussion.
This goes for the principles corresponding to accuracy - loging mistakes is good practice no matter where -, auditability - one should never obstruct uninterested parties' inquiries into the possibility of things going wrong - and explainability - obfuscating one's work and keeping affected parties ignorant is evidently bad practice both in ML and in DS.
Much more interesting than the overlap lately mentioned, however, is the question how the two remaining principles apply to to DS.

@@

In this fashion fairness principle could be addressed in DS. This does not yet respond, however, to the question where exactly responsibility lies - that is, with whom - in case something does go wrong.
It is perhaps the most intricate difficulty concerning ethics in DS as potential culprits are society at large, the producers of the corpus, the distributional semanticist or the person who implements the system that interacts with society again. 
The nature of language meaning and how language and society interact are notoriously contested - the heated debates about what is appropriate in gendering language serves as a case in point. For data engineers and programmers who work with these systems there is certainly no easy answer to this issue at first sight.
The best that can be offered in terms of practical advice is that if the possibilty exists to algorithmically rid a DS model of biases relevant to its application, then this should be done.
\section{Conclusion}\hypertarget{sec5}{}

\section{References}\hypertarget{sec6}{ }
\printbibliography

>>>>

We will now briefly go over the five principles - \emph{responsibility, explainability, accuracy, auditability and fairness} to describe their original intention in the PAA.
Responsibility requires developers to take seriously the possibility that adverse effects are produced their algorithmic system. Consequently, there should be designated roles within the organization to be responsible for redress if harm is caused. Excuses akin to ''the algorithm did it'' are no valid scapegoats. This reflects the acknowledgment that, at some level, there are people who design these systems and put them to use.
Explainability demands that both data selection and decisions taken by the algorithms can be explained to people affected by its usage in terms that are understandable by laywomen. The point here is not to obscure the workings of the systems to a point where no stakeholder can grasp it.
Accuracy asks designers and users of the system to keep track of its mistakes and sources of error, so that worst case scenarios become more easily anticipated and possible damage reduced.
Audibility is the requirement to create transparency in such a way that interested third parties are in the position to review the algorithms performance. This point is related to explainability.
Finally, Fairness requires that the algorithmic system does not affect different demographic groups in disparate and unfavorable ways, that is, discriminates against them.
We now want to assess to what degree we can use these principles for the field of DS.
\section{Principles for Distributional Semantics}\hypertarget{sec4}{}
It should be fair to say that some of these principles allow seamless transfer from general ML to any specific sub-domain, as DS, without much further discussion.
This goes for the principles corresponding to accuracy - logging mistakes is good practice no matter where -, auditability - one should never obstruct uninterested parties' inquiries into the possibility of things going wrong - and explainability - obfuscating one's work and keeping affected parties ignorant is evidently bad practice both in ML and in DS.
Much more interesting than the overlap lately mentioned, however, is the question how the two remaining principles apply to to DS.

@@

In this fashion fairness principle could be addressed in DS. This does not yet respond, however, to the question where exactly responsibility lies - that is, with whom - in case something does go wrong.
It is perhaps the most intricate difficulty concerning ethics in DS as potential culprits are society at large, the producers of the corpus, the distributional semanticist or the person who implements the system that interacts with society again. 
The nature of language meaning and how language and society interact are notoriously contested - the heated debates about what is appropriate in gendering language serves as a case in point. For data engineers and programmers who work with these systems there is certainly no easy answer to this issue at first sight.
The best that can be offered in terms of practical advice is that if the possibilty exists to algorithmically rid a DS model of biases relevant to its application, then this should be done.
\section{Conclusion}\hypertarget{sec5}{}
The preceding discussion seems to warrant but two clear take-aways for DS. First, if a semantic space does contain bias and will be further used in some application, then efforts should be undertaken to debias it. Second, if no the DS model cannot be debiased, then it should not be employed externally without further efforts to ensure that it does not discriminate. The responsibility for these two points ultimately lie with the designer of the application.
\section{References}\hypertarget{sec6}{ }
\printbibliography
